---
title: "Clinical codes - lookups and mappings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical codes - lookups and mappings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ukbwranglr)
library(magrittr)
```

# Introduction

This vignette describes how to work with clinical codes using `ukbwranglr`. Specifically:

- Looking up codes and their descriptions
- Getting 'children' codes
- Mapping codes from one coding system to another

The functions provided by `ukbwranglr` rely on [UK Biobank resource 592](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592), which includes an Excel workbook containing lookup and mapping tables. This has been converted into a named list of data frames, where each item in the list corresponds to a sheet in the original Excel workbook. This can be retrieved with `get_ukb_code_mappings()`:

```{r}
# retrieve code mappings in .Rdata format
ukb_code_mappings <- get_ukb_code_mappings()

# each item in the list is a sheet in the UKB Excel workbook (resource 592)
names(ukb_code_mappings)
```

> **Note:** `get_ukb_code_mappings()` returns a large object (~460MB)

# Code lookups

To lookup details for a particular code, use `lookup_codes()`. Setting `preferred_description_only` to `TRUE` will return only the preferred code descriptions if synonyms are present (`read2` and `read3` may include multiple descriptions for the same code):

```{r}
# Some Read2 codes for T1DM
t1dm_read2 <- c("C10E.", "C108.")

# lookup details
lookup_codes(codes = t1dm_read2,
             code_type = "read2",
             ukb_code_mappings = ukb_code_mappings,
             preferred_description_only = TRUE)
```
By default, the output is standardised to produce the columns shown above. To output the original formatting from [UK Biobank resource 592](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592), set `standardise_output` to `FALSE`:

```{r}
# lookup details
lookup_codes(codes = t1dm_read2,
             code_type = "read2",
             ukb_code_mappings = ukb_code_mappings,
             preferred_description_only = TRUE,
             standardise_output = FALSE)
```

# Code mapping

An example: map codes for type 1 diabetes from Read2 to Read3. Note that both of the Read 2 codes `C10E.` and `C108.` map to a single read3 code, `X40J4`:

```{r}
# Some Read2 codes for T1DM
t1dm_read2 <- c("C10E.", "C108.")

# lookup details
map_codes(codes = t1dm_read2, 
          from = "read2", 
          to = "read3", 
          ukb_code_mappings = ukb_code_mappings)
```

Note that `preferred_description_only` cannot be `TRUE` if `standardise_output` is `FALSE` with `map_codes` (will raise an error). This is because some codes may otherwise be 'lost' in the mapping process. When `standardise_output` is `TRUE`, the mapped codes from `map_codes` are passed on to `lookup_codes`, at which point one can request to return only preferred code descriptions:

```{r}
# mapping the Read 2 code "D4104" to Read 3 only returns the secondary Read 3
# description (`TERMV3_TYPE` == "S"), unlike for "D4103". 
map_codes(codes = c("D4103", "D4104"), 
          from = "read2", 
          to = "read3", 
          ukb_code_mappings = ukb_code_mappings, 
          codes_only = FALSE,
          preferred_description_only = FALSE,
          standardise_output = FALSE) %>% 
  dplyr::select(tidyselect::contains("V3"))
```

```{r}
# if `standardise_output` is `TRUE`, then `preferred_description_only` may also
# be set to `TRUE`
map_codes(codes = c("D4103", "D4104"), 
          from = "read2", 
          to = "read3", 
          ukb_code_mappings = ukb_code_mappings, 
          codes_only = FALSE,
          preferred_description_only = TRUE,
          standardise_output = TRUE)
```

Mapping to ICD is more problematic as some results are a range of ICD codes (note also that for ICD-10, the mapping sheets use an alternative code format which removes any "." characters):

```{r}
map_codes(codes = t1dm_read2, 
          from = "read2", 
          to = "icd10", 
          ukb_code_mappings = ukb_code_mappings, 
          codes_only = TRUE,
          standardise_output = FALSE)
```

The available mappings do not cover all possible mapping directions. For example, while there are mappings for Read2 to ICD-10, there is no mapping for ICD-10 to Read2. For cases like this, `map_codes()` will attempt to map anyway by using the same mapping sheet in reverse (e.g. for mapping ICD-10 to Read2, `map_codes` uses the `read_v2_icd10` mapping sheet). However, this returns no results when attempting to map the ICD-10 code, `H36.0` for diabetic retinopathy:

```{r}
# find ICD-10 code matching "diabetic retinopathy"
icd10_diabetic_retinopathy <-
  search_codes_by_description(
    reg_expr = "diabetic retinopathy",
    code_type = "icd10",
    ukb_code_mappings = ukb_code_mappings,
    ignore_case = TRUE,
    codes_only = TRUE
  )

# attempting to map this to Read 2 returns a NULL result however
map_codes(
  codes = icd10_diabetic_retinopathy,
  from = "icd10",
  to = "read2",
  ukb_code_mappings = ukb_code_mappings,
  standardise_output = FALSE,
  codes_only = TRUE
)
```

Inspecting the mapping sheet `read_v2_icd10` shows why. The `icd10_code` column contains 2 ICD-10 codes which both describe diabetic retinopathy:

```{r}
ukb_code_mappings$read_v2_icd10 %>% 
  dplyr::filter(stringr::str_detect(icd10_code, pattern = "H360"))
```

> **Note:** "H36.0" is converted to "H360" by `map_codes` internally, as this is the format used by the mapping sheets.

# Find codes that match a description

Use `search_codes_by_description()`. For example, to find Read2 codes that match the description 'diabetic retinopathy':

```{r}
search_codes_by_description(reg_expr = "diabetic retinopathy", 
                          code_type = "read2", 
                          ukb_code_mappings = ukb_code_mappings, 
                          ignore_case = TRUE, 
                          codes_only = FALSE,
                          preferred_description_only = TRUE) %>% 
  head()
```

# Children codes

To get the children codes, use `get_child_codes()`. This will return all unique clinical codes that start with the codes of interest:

```{r}
# Some Read2 codes for T1DM
t1dm_read2 <- c("C10E.", "C108.")

# lookup details
get_child_codes(codes = t1dm_read2, 
                code_type = "read2", 
                ukb_code_mappings = ukb_code_mappings, 
                codes_only = TRUE)
```

> **Note:** Some coding systems include a '.' (e.g. ICD-10) - this may return unexpected results with `get_child_codes()`, as this function searches using regexs and '.' is interpreted as a wildcard.

By default, a character vector of codes is returned. To return a data frame including code descriptions, set the argument `codes_only` to `FALSE`:

```{r}
# Some Read2 codes for T1DM
t1dm_read2 <- c("C10E.", "C108.")

# lookup details
get_child_codes(codes = t1dm_read2, 
                code_type = "read2", 
                ukb_code_mappings = ukb_code_mappings, 
                codes_only = FALSE) %>% 
  head()
```
