---
title: "Clinical codes - lookups and mappings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical codes - lookups and mappings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ukbwranglr)
library(magrittr)
```

# Introduction

This vignette describes how to work with clinical codes using `ukbwranglr`. Specifically:

- Looking up codes and their descriptions
- Getting 'children' codes
- Mapping codes from one coding system to another

The functions provided by `ukbwranglr` rely on [UK Biobank resource 592](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=592), which includes an Excel workbook containing lookup and mapping tables. This has been converted into a named list of data frames, where each item in the list corresponds to a sheet in the original Excel workbook. This can be retrieved with `get_ukb_code_mappings()`:

```{r}
# retrieve code mappings in .Rdata format
ukb_code_mappings <- get_ukb_code_mappings()

# each item in the list is a sheet in the UKB Excel workbook (resource 592)
names(ukb_code_mappings)
```

# Code lookups

To lookup details for a particular code, use `lookup_codes()`:

```{r}
# A Read2 code for T1DM
t1dm_read2 <- c("C10E.", "C108.")

# lookup details
lookup_codes(codes = t1dm_read2,
             code_type = "read2",
             ukb_code_mappings = ukb_code_mappings)
```

# Children codes

To get the children codes, use `get_child_codes()`. This will return all unique clinical codes that start with the codes of interest:

```{r}
get_child_codes(codes = t1dm_read2, 
                code_type = "read2", 
                ukb_code_mappings = ukb_code_mappings, 
                codes_only = TRUE)
```

> **Note:** Some coding systems include a '.' (e.g. ICD-10) - this may return unexpected results with `get_child_codes()`, as this function searches using regexs and '.' is interpreted as a wildcard.

By default, a character vector of codes is returned. To return a data frame including code descriptions, set the argument `codes_only` to `FALSE`:

```{r}
get_child_codes(codes = t1dm_read2, 
                code_type = "read2", 
                ukb_code_mappings = ukb_code_mappings, 
                codes_only = FALSE) %>% 
  head()
```

# Find codes that match a description

Use `search_codes_by_description()`. For example, to find Read2 codes that match the description 'diabetic retinopathy':

```{r}
search_codes_by_description(reg_expr = "diabetic retinopathy", 
                          code_type = "read2", 
                          ukb_code_mappings = ukb_code_mappings, 
                          ignore_case = TRUE, 
                          codes_only = FALSE) %>% 
  head()
```

# Code mapping

Map codes for type 1 diabetes from Read2 to Read3:

```{r}
map_codes(codes = t1dm_read2, 
          from = "read2", 
          to = "read3", 
          ukb_code_mappings = ukb_code_mappings, 
          codes_only = TRUE)
```

Mapping to ICD is more problematic as some results are a range of ICD codes (note also that for ICD-10, the mapping sheets use an alternative code format which removes any "." characters):

```{r}
map_codes(codes = t1dm_read2, 
          from = "read2", 
          to = "icd10", 
          ukb_code_mappings = ukb_code_mappings, 
          codes_only = TRUE)
```

The available mappings do not cover all possible mapping directions. For example, while there are mappings for Read2 to ICD-10, there is no mapping for ICD-10 to Read2. For cases like this, `map_codes()` will attempt to map anyway:

```{r}
# find ICD-10 code matching "diabetic retinopathy"
search_codes_by_description(
  reg_expr = "diabetic retinopathy",
  code_type = "icd10",
  ukb_code_mappings = ukb_code_mappings,
  ignore_case = TRUE,
  codes_only = FALSE
)
```

However, this returns a `NULL` result when attempting to map the following:

```{r}
# try mapping this ICD-10 code for diabetic retinopathy to Read codes
icd10_diabetic_retinopathy <-
  search_codes_by_description(
    reg_expr = "diabetic retinopathy",
    code_type = "icd10",
    ukb_code_mappings = ukb_code_mappings,
    ignore_case = TRUE,
    codes_only = TRUE
  )

# this returns a `NULL` result however
icd10_diabetic_retinopathy_as_read_code <- map_codes(
  codes = icd10_diabetic_retinopathy,
  from = "icd10",
  to = "read2",
  ukb_code_mappings = ukb_code_mappings,
  codes_only = TRUE
)
```

Inspecting the mapping sheet `read_v2_icd10` shows why:

> **Note:** "H36.0" is converted to "H360" by `map_codes` internally, as this is the format used by the mapping sheets.

```{r}
ukb_code_mappings$read_v2_icd10 %>% 
  dplyr::filter(stringr::str_detect(icd10_code, pattern = "H360"))
```

