---
title: 'UKB: Disease outcomes'
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(ukbwranglr)
library(tidyverse)
library(readr)
library(configr)

# Constants
config <- read.config("../config.ini")
DUMMY_DATA_PATH <- paste0("../", config$PATHS$DUMMY_DATA_PATH)
GP_CLINICAL_AK <- config$PATHS$GP_CLINICAL_AK
GP_CLINICAL_AK_TEMP <- config$PATHS$GP_CLINICAL_AK_TEMP
UKB_DB <- config$PATHS$UKB_DB

# Load data
dummy_data_dict <- ukbwranglr::pheno_data_dict(DUMMY_DATA_PATH, delim = ",")
dummy_ukb_data <- ukbwranglr::read_pheno(DUMMY_DATA_PATH, dummy_data_dict, delim = ",") 
dummy_ukb_data$eid <- as.character(seq(1:nrow(dummy_ukb_data)))
ukb_codings <- data.table::fread(
    "https://biobank.ctsu.ox.ac.uk/~bbdatan/Codings.tsv",
    colClasses = c('character'),
    sep = "\t",
    quote = " ",
    na.strings = c("", "NA")
  )

dummy_ukb_data_raw <- data.table::fread(
    DUMMY_DATA_PATH,
    colClasses = c('character'),
    sep = ",",
    quote = " ",
    na.strings = c("", "NA")
  )

# ukb_data_dict <- data.table::fread(
#     "https://biobank.ctsu.ox.ac.uk/~bbdatan/Data_Dictionary_Showcase.tsv",
#     colClasses = c('character'),
#     na.strings = c("", "NA"),
#     sep = "\t",
#   )
```

# Overview

Disease outcomes in UKB may be primarily ascertained from the following sources:

- [Self-report data](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=100074): [cancer](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20001) and [non-cancer](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20002) 
- [Hospital inpatient data](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=2000)
- [Death register records](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=100093)
- [Cancer register data](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=100092)
- [Primary care data](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=3000)

>**Note:** the preceding links for self-report data relate to verbal interview responses. Self-report data also includes [touchscreen](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=100025) responses, however UKB [resource 596](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=596) recommends that the verbal interview data fields are used in preference

UKB also provides a set of '[first occurrences](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=1712)' fields, which record the earliest date a 3-character ICD-10 code (e.g. ICD-10 code 'E11' = 'Non-insulin-dependent diabetes mellitus') appears in one of the categories listed above. **Note: this does not include cancer codes, as these are comprehensively captured by the [cancer register data](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=100092)**

Other sources to consider include: 

- '[Algorithmically-defined outcomes](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=42)'
- Operations (self-reported; codes are otherwise included in the data sources considered here)
- Medication history (e.g. classify participants as having diabetes if they take insulin) - see `vignette("ukb_medications")`

The rest of this article details how disease outcomes may be extracted from each of the data sources outlined above using `ukbwranglr`.

# Self-report data ([verbal interview](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=100074))

## Introduction

FieldID's for self-reported (verbal interview) disease outcomes are categorised into 'non-cancer' and 'cancer':

| Field | Non-cancer | Cancer |
|:------|------------|--------|
| Illness code, self-reported | [20002](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20002) ([data-coding 6](https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=6)) | [20001](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20001) ( [data-coding 3](https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=3))|
| Interpolated year when first diagnosed | [20008](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20008) | [20006](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20006) |
| Interpolated age when first diagnosed | [20009](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20009) | [20007](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=20007) |

UK Biobank also provides mappings from self-reported non-cancer illness codes to 3-character ICD-10 codes ([data-coding 609](https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=609)). 

>**NOTE:** Not all the self-report codes are mapped to ICD-10 (111 out of 447). For example, the non-cancer illness code `1065` refers to 'hypertension', which cannot be mapped to ICD-10 as it is too non-specific Code `1072`, which refers to 'essential hypertension' however *is* mappable (maps to ICD-10 code `I10`). 

```{r eval=FALSE, include = FALSE}
# Self-report codes that do not map to ICD-10 may be identified as follows:

# get UKB codings dictionary
# (https://biobank.ctsu.ox.ac.uk/crystal/exinfo.cgi?src=accessing_data_guide),
# filtered for coding 609
ukb_codings_609 <-
  readr::read_tsv(
    "https://biobank.ctsu.ox.ac.uk/~bbdatan/Codings.tsv",
    col_types = cols(.default = "c"),
    quote = ""
  ) %>%
  filter(Coding == "609")

# get the full set of UKB self-report non-cancer illness codes form website
# https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=6
html_page <-
  xml2::read_html("https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=6")
table_nodes <- xml2::xml_find_all(html_page, "//table")
ukb_selfreport_non_cancer_codings_readable <-
  rvest::html_table(table_nodes[[2]])

# filter for those that were selectable (i.e. no 'special' values like -1)
ukb_selfreport_non_cancer_codings_readable <-
  ukb_selfreport_non_cancer_codings_readable %>%
  filter(Selectable == "Yes")

# filter for those NOT mapped to ICD-10
ukb_selfreport_non_cancer_codings_readable <-
  ukb_selfreport_non_cancer_codings_readable %>%
  filter(!(Coding %in% ukb_codings_609$Value))

View(ukb_selfreport_codings_readable_no_special_values_not_in_ukb_codings_609)
```

```{r eval=FALSE, include=FALSE}
# ***EXPLORE***

# Note: the ukb codings-file contains all the codes on the webpage for data-coding 609 https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=609
# get ukbdata-coding 609 (map self-report codes to icd-10 codes)
html_page <- xml2::read_html("https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=609")
table_nodes <- xml2::xml_find_all(html_page, "//table")
ukb_selfreport_icd_map <- rvest::html_table(table_nodes[[2]])

# ...returns TRUE
all(ukb_codings_609$Value %in% ukb_selfreport_icd_map$Coding)
```

>**NOTE:** Self-reported cancer conditions are _not_ mapped to ICD10 as they are comprehensively captured by the cancer register. They are also _not_ included in the 'first occurrences' data fields (see [resource 593](https://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=593))

## Identifying health outcomes using diagnostic codes

### Non-cancer medical conditions

**Step 1:** A long format table summarising all self-reported non-cancer medical conditions (coded as per [data coding 6](https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=6)) for each UK Biobank participant can be created as follows:

```{r eval = FALSE}
# get UKB codings file
ukb_codings <- data.table::fread(
    "https://biobank.ctsu.ox.ac.uk/~bbdatan/Codings.tsv",
    colClasses = c('character'),
    sep = "\t",
    quote = " ",
    na.strings = c("", "NA")
  )

self_report_non_cancer_diagnoses <- get_self_report_non_cancer_diagnoses(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)
```

Alternatively, this can be created using [ICD10 code mappings](https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=609) (see note above):

```{r eval = FALSE}
self_report_non_cancer_diagnoses_icd10 <- get_self_report_non_cancer_diagnoses_icd10(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)
```

```{r include = FALSE}
self_report_non_cancer_diagnoses_icd10 <- get_self_report_non_cancer_diagnoses_icd10(
  ukb_pheno = dummy_ukb_data, 
  data_dict = dummy_data_dict, 
  ukb_codings = ukb_codings
)
```

**Step 2:** These may then be used to retrieve UK Biobank participants with one or more specified codes in these records, as well as the earliest recorded date (i.e. the earliest date for any one of these codes):

```{r}
# select ICD10 codes
selected_codes <- c("A014", "N12", "L86")

# extract eids with these codes, as well as the earliest recorded date
get_first_diagnostic_code_record(self_report_non_cancer_diagnoses_icd10,
                                 selected_codes) %>% 
  head() # display first few rows only
```

### Cancer

Follow the same work flow as for non-cancer medical conditions, using `get_self_report_cancer_diagnoses` instead for step 1.

```{r eval=FALSE, include=FALSE}
self_report_cancer_codes <- as.character(1000:1030)

self_report_cancer_diagnoses <- get_self_report_cancer_diagnoses(
  ukb_pheno = dummy_ukb_data, 
  data_dict = dummy_data_dict, 
  ukb_codings = ukb_codings
)

# extract eids with these codes, as well as the earliest recorded date
get_first_diagnostic_code_record(self_report_cancer_diagnoses,
                                 self_report_cancer_codes) %>% 
  head() # display first few rows only
```

# Hospital inpatient data - summary diagnoses

## Introduction

These fields ([category 2002](https://biobank.ndph.ox.ac.uk/ukb/label.cgi?id=2002)) summarise diagnoses made during hospital inpatient admissions, recorded as either ICD9 or ICD10 codes: 

| Field | ICD9 | ICD10 |
|:------|------|-------|
| Diagnoses | [41271](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=41271) ([data-coding 87](https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=87)) | [41270](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=41270) ([data-coding 19](https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=19))|
| Date of first in-patient diagnosis | [41281](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=41281) | [41280](https://biobank.ndph.ox.ac.uk/ukb/field.cgi?id=41280) |

>**Note:** Diagnoses are categorised into main and secondary diagnoses, with separate FieldIDs provided for both (further subdivided into ICD-9 and ICD-10). The table above lists those summary fields which include both main and secondary diagnoses.

*See also [resource 141140](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=141140) for further details re hospital inpatient data*

## Identifying health outcomes using diagnostic codes

Use the same work flow as described for self-report data. To generate a long format table summarising all self-reported non-cancer medical conditions for each UK Biobank participant, use `get_hes_icd10_diagnoses()` or `get_hes_icd9_diagnoses()`:

```{r eval = FALSE}
hes_icd10_diagnoses <- get_hes_icd10_diagnoses(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)
```

```{r include = FALSE}
hes_icd10_diagnoses <- get_hes_icd10_diagnoses(
  ukb_pheno = dummy_ukb_data, 
  data_dict = dummy_data_dict, 
  ukb_codings =  ukb_codings
)
```

Then use `get_first_diagnostic_code_record()`:

```{r}
# select ICD10 codes
selected_codes <- c("A014", "N12", "L86")

# extract eids with these codes, as well as the earliest recorded date
get_first_diagnostic_code_record(hes_icd10_diagnoses,
                                 selected_codes) %>%
  head() # display first few rows only
```

# Death register records

## Introduction

Cause of death data [category 100093](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=100093) is available as part of a main UK Biobank dataset as ICD10 codes under FieldIDs [40001](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40001) (primary cause of death) and [40002](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40002) (secondary causes of death).

See also [resource 115559](https://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=115559) for more information.

## Identifying health outcomes using diagnostic codes

Use the same work flow as described for self-report data. To generate a long format table summarising all self-reported non-cancer medical conditions for each UK Biobank participant, use `get_death_data_icd10_diagnoses()`:

```{r eval = FALSE}
death_data_icd10_diagnoses <- get_death_data_icd10_diagnoses(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)
```

```{r include = FALSE}
death_data_icd10_diagnoses <- get_death_data_icd10_diagnoses(
  ukb_pheno = dummy_ukb_data, 
  data_dict = dummy_data_dict, 
  ukb_codings = ukb_codings
)
```

Then use `get_first_diagnostic_code_record()`:

```{r}
# select ICD10 codes
selected_codes <- c("A014", "N12", "L86", "M7704")

# extract eids with these codes, as well as the earliest recorded date
get_first_diagnostic_code_record(death_data_icd10_diagnoses,
                                 selected_codes) %>%
  head() # display first few rows only
```

# Cancer register records

## Introduction

Coded data on cancer diagnoses ([category 100092](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=100092)) obtained through linkage to the national cancer registries are available from the following fields:

| FieldID | Description |
|---------|:------------|
| [40013](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40013) | Type of cancer: ICD9 |
| [40006](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40006) | Type of cancer: ICD10 |
| [40008](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40008) | Age at cancer diagnosis |
| [40005](https://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40005) | Date of cancer diagnosis |

*See also [resource 115558](https://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=115558) for further details re cancer register data*

## Identifying health outcomes using diagnostic codes

Use the same work flow as described for self-report data. To generate a long format tables (ICD-9 and ICD-10) summarising all cancer conditions recorded for each UK Biobank participant, use `get_cancer_register_icd9_diagnoses()` and `get_cancer_register_icd10_diagnoses()`:

```{r eval=FALSE}
cancer_register_icd9_diagnoses <- get_cancer_register_icd9_diagnoses(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)

cancer_register_icd10_diagnoses <- get_cancer_register_icd10_diagnoses(
  ukb_pheno, # dataset read into R using read_pheno()
  data_dict, # data dictionary created with pheno_data_dict()
  ukb_codings # the UKB codings file
)
```

```{r include=FALSE}
cancer_register_icd9_diagnoses <- get_cancer_register_icd9_diagnoses(
  ukb_pheno = dummy_ukb_data,
  data_dict = dummy_data_dict,
  ukb_codings = ukb_codings
)

cancer_register_icd10_diagnoses <- get_cancer_register_icd10_diagnoses(
  ukb_pheno = dummy_ukb_data, 
  data_dict = dummy_data_dict, 
  ukb_codings = ukb_codings
)
```

Then use `get_first_diagnostic_code_record()`:

```{r}
# select ICD10 codes
selected_codes <- c("A001", "A156", "E111")

# extract eids with these codes, as well as the earliest recorded date
get_first_diagnostic_code_record(cancer_register_icd10_diagnoses,
                                 selected_codes) %>%
  head() # display first few rows only
```

# Ascertaining disease outcomes from multiple sources

Supply a list of functions from the 'get all diagnostic codes' family to `get_all_diagnostic_codes_multi()` (see function documentation for defaults), then apply `get_first_diagnostic_code_record()` to the result:

```{r eval=FALSE}
# TEMP
test <- list(
    get_death_data_icd10_diagnoses,
    get_hes_icd9_diagnoses,
    get_hes_icd10_diagnoses,
    get_self_report_non_cancer_diagnoses,
    get_self_report_non_cancer_diagnoses_icd10,
    get_self_report_cancer_diagnoses,
    get_cancer_register_icd9_diagnoses,
    get_cancer_register_icd10_diagnoses
  ) %>% 
  map(~ get_all_diagnostic_codes_multi(
      ukb_pheno = dummy_ukb_data,
      data_dict = dummy_data_dict,
      ukb_codings = ukb_codings,
      function_list = list(.x)))

test2 <- bind_rows(test)
```


```{r}
# list of functions
list_of_get_diagnostic_codes_functions <-
  list(
    get_self_report_non_cancer_diagnoses,
    get_self_report_non_cancer_diagnoses_icd10,
    get_hes_icd10_diagnoses,
    get_cancer_register_icd10_diagnoses
  )

# select ICD10 codes
selected_codes <- c("A001", "A156", "E111")

# get all diagnostic codes for selected data sources
diagnostic_codes_multiple_sources <-
  get_all_diagnostic_codes_multi(ukb_pheno = dummy_ukb_data,
                                 data_dict = dummy_data_dict,
                                 ukb_codings = ukb_codings,
                                 function_list = list_of_get_diagnostic_codes_functions)

# extract eids with these codes, as well as the earliest recorded date (display
# first few rows only)
get_first_diagnostic_code_record(diagnostic_codes_multiple_sources,
                                 selected_codes) %>%
  head()
```

# Primary care data

## Introduction

The interim release primary care dataset covers ~50% of the full UK Biobank cohort and is provided in 3 tables:

1. Clinical events
2. Prescriptions
3. Registrations

Only the clinical events table will be considered here. This consists of >120 million rows with the following columns:

- `eid`: unique participant identifier
- `event_dt`: record date
- `read_2`: Read 2 code
- `read_2`: Read 3 code
- `value_1`, `value_2` and `value_3`: free text values (e.g. blood pressure measurements)

*See resources 591 and 592 for further information (available from [here](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=3000)).*

## Identifying health outcomes using diagnostic codes

**Step 1:** Write the UKB primary care clinical events dataset to a table in a SQLite database (the database will be created if it does not already exist). The `file_to_sqlite_db()` function reads a file in chunks (chunk size as specified by the `chunk_size`), pre-processes the 'chunk' if the `data_processing_function` is not set to `NULL`, then writes to a SQLite database. The `gp_clinical_to_sqlite_db` function reshapes and formats the data to match output from `get_all_diagnostic_codes_multi`. By default, special date values are set to `NA` (see [resource 591](https://biobank.ndph.ox.ac.uk/ukb/refer.cgi?id=591) for further details).

The interim release primary care clinical events file is ~4GB large with >120 million rows. Writing to a database may take ~5 minutes.

```{r eval=FALSE}
# create a database connection (this will create a new database if it does not already exist)
con = DBI::dbConnect(RSQLite::SQLite(), "ukb.db")

# write to table
file_to_sqlite_db(file = "PATH_TO_GP_CLINICAL.TXT", 
             db_path = "PATH_TO_DATABASE",
             table = "clinical_events",
             chunk_size = 500000,
             delim = "\t",
             append = FALSE,
             verbose = TRUE,
             data_processing_function = gp_clinical_to_sqlite_db,
             remove_special_dates = TRUE) 
```


```{r eval=FALSE, include=FALSE}
# connect to db
con = DBI::dbConnect(RSQLite::SQLite(), UKB_DB)

# write file to table named 'gp_clinical' in database
file_to_sqlite_db(file = GP_CLINICAL_AK_TEMP, 
             col_types = readr::cols(.default = "c"),
             db_path = UKB_DB,
             table = "clinical_events",
             chunk_size = 500000,
             delim = "\t",
             append = FALSE, # set to TRUE if appending to an existing table
             verbose = TRUE,
             data_processing_function = gp_clinical_to_sqlite_db,
             remove_special_dates = TRUE)
```

**Step 2:** Filter for selected codes - retrieve the table using `dplyr::tbl()`, `dplyr::filter()` for selected codes, read the results into R with `dplyr::collect()` and convert the 'date' column to date format:

```{r eval=FALSE}
# retrieve table
clinical_events <- dplyr::tbl(con, "clinical_events") 

# filter for selected codes
selected_codes <- c("Xa8Pl", "246A.")
eids_with_selected_codes <- clinical_events %>% 
  filter(code %in% selected_codes) %>% 
  collect() %>% 
  mutate(date = as.Date(date))
```

**Step 3:** Then use `get_first_diagnostic_code_record` to identify participants with specific codes of interest in their record. See the [dbplyr getting started vignette](https://dbplyr.tidyverse.org/articles/dbplyr.html) for a good overview of querying databases using `dplyr` syntax.

```{r eval=FALSE, include=FALSE}
get_first_diagnostic_code_record(eids_with_selected_codes, 
                                 selected_codes)
```

# Combining all diagnostic codes into a SQLite database

**Step 1:** Follow 'step 1' in 'Primary care data', if not already done.

**Step 2:** Append to this the output from get_all_diagnostic_codes_multi (set the `append` argument in `dbWriteTable` to `TRUE`).

```{r eval=FALSE}
# get diagnostic codes from ukb pheno data
diagnostic_codes_multiple_sources <-
  get_all_diagnostic_codes_multi(ukb_pheno = dummy_ukb_data,
                                 data_dict = dummy_data_dict,
                                 ukb_codings = ukb_codings)

# append to `clinical_events` table
main_dataset_diagnoses_to_sqlite_db(diagnostic_codes_multiple_sources,
                                                "clinical_events",
                                                db_path = "PATH_TO_DATABASE",
                                                overwrite = FALSE,
                                                append = TRUE)

main_dataset_diagnoses_to_sqlite_db(diagnostic_codes_multiple_sources,
                                                "clinical_events",
                                                db_path = "ukb.db",
                                                overwrite = FALSE,
                                                append = TRUE)
```

**Step 3:** Then use `get_first_diagnostic_code_record_db` to identify participants with specific codes of interest in their record. See the [dbplyr getting started vignette](https://dbplyr.tidyverse.org/articles/dbplyr.html) for a good overview of querying databases using `dplyr` syntax.

>**NOTE:** To improve search performance, set an index on the 'code' column (this will be filtered on most frequently) e.g. `dplyr::db_create_index(con, "clinical_events", "code")`.

>**NOTE:** To remove a table, use `DBI::dbRemoveTable(conn, 'table_name')`.
